Income Disparity and Its Influence on Higher Education Affordability and Access\thanks{Code and data are available at: https://github.com/fanger2791/College-Tuition-and-Income-Inequality/tree/main}

### 2024-02-20 files/paper/paper.tex
update

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

\title{Income Disparity and Its Influence on Higher Education
Affordability and Access\thanks{Code and data are available at:
https://github.com/fanger2791/College-Tuition-and-Income-Inequality/tree/main.
Replication on Social Science Reproduction:
https://www.socialsciencereproduction.org/reproductions/d456357c-d73d-4bce-a6ba-178bac339c7e/index}}
\author{Michael Fang \and Harrison Huang}
\date{February 19, 2024}

@@

In recent years, the escalating cost of college tuition in the United
States has emerged as a critical issue, intertwining with the broader
narrative of rising income inequality. The paper by Zhifeng Cai and
Jonathan Heathcote addresses this pressing concern by examining the
relationship between these two phenomena. Utilizing a competitive model
of the college market, the study not only uncovers the mechanisms that
link rising income inequality to increasing tuition costs but also sheds
light on the consequent impacts on college attendance and social
mobility (Cai and Heathcote 2022). Through this paper, we aim to
illuminate the unique challenges and opportunities facing the Canadian
higher education system in the context of rising tuition costs and
income inequality. Incorporating a Canadian perspective into our
reproduction study adds a crucial comparative dimension, allowing us to
explore the dynamics of university tuition and income inequality within
a different national context. This extension acknowledges the distinct
structure of the Canadian higher education system and its funding
mechanisms, which differ from those in the United States in terms of
government support, the magnitude of tuition fees, and the
socio-economic landscape.

Central to the Zhifeng Cai and Jonathan Heathcote paper's findings is
the assertion that the surge in U.S. income inequality since the 1990s
can account for more than half of the observed increase in average net
tuition over the same period (Cai and Heathcote 2022). This relationship
underscores a critical feedback loop where income disparities not only
affect individual capacity to afford higher education but also drive
institutional behaviors around tuition setting. Moreover, the paper
highlights the detrimental effects of rising tuition on college
attendance rates, particularly among lower-income segments, positing
significant implications for social mobility and equity (Cai and
Heathcote 2022).

@@

Our project involves crafting a concise paper that encapsulates the
entire reproduction study. This paper will meticulously outline our
source and methodology, ensuring clarity on how we've mirrored the
original study's model while integrating steps for high reproducibility.
It will also present a comparative analysis of our findings against the
original study's conclusions, pinpointing any variances and their
potential causes. Additionally, the paper will delve into the
implications of our policy simulations, offering insights into how
income inequality influences college tuition and attendance. Lastly, it
will propose directions for future research, highlighting areas within
the college tuition-income inequality that remain unexplored, setting
the stage for subsequent scholarly inquiry.

@@

This paper reproduces several areas of the original paper by Zhifeng Cai
and Jonathan Heathcote (Cai and Heathcote 2022) which is originally from
the American Economic Association. This reproduction utilizes the same
data sets in the process. Please note that the datasets used are
unfortunately not publicly accessible.

@@

In replicating the original model by Zhifeng Cai and Jonathan Heathcote,
our methodology was crafted to mirror their approach with precision,
ensuring the highest degree of reproducibility. This entailed a thorough
analysis of the original study's model, including its assumptions,
variables, and computational methods. We closely followed the procedural
steps outlined in their research, from data collection through to the
analytical techniques employed to examine the relationship between
income inequality and college tuition costs. To guarantee
reproducibility, we documented each step of our process in detail,
including the coding practices, statistical software used, and the
sources of our data. This approach not only underscores our commitment
to transparency and scientific integrity but also enhances the
reliability of our replication efforts.

R (R Core Team 2023) was the language and environment used for this
reproduction, with different packages such as tidyverse (Wickham et al.
2019), ggplot2 (Wickham 2016), readxl (Wickham and Bryan 2023), dplyr
(Wickham et al. 2023), tidyr (Wickham et al. 2024), knitr (Xie 2023)
packages.

\subsection{Variables}\label{variables}

Variables within the datasets have been constructed to reflect the
nuances of income distribution across different socioeconomic groups and
the corresponding tuition rates at a range of institutions, from
community colleges to elite universities.

We have listed the different variables used below and how they
correspond:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Net Prices Data:
\end{enumerate}

public\_college\_netprice: Data on the net prices of public colleges.

private\_college\_netprice: Data on the net prices of private colleges.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  Grants Data:
\end{enumerate}

public\_college\_grants: Data on total grants by income and year for
public four-year colleges.

private\_college\_grants: Similar data for private four-year colleges.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  Funding and Revenue Data:
\end{enumerate}

state\_funding: average state and local funding for higher education per
1000\$ of personal income

public\_rev\_subs and private\_rev\_subs: detailing subsidies and
revenues for public and private four-year institutions, with columns for
institution type, year, net tuition revenue, subsidy, and
education-related expenditures.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\tightlist
\item
  College Tuition and Fees Data:
\end{enumerate}

Public Sticker: This is the advertised or sticker price for attending a
public college or university. It includes tuition, fees, room and board,
and other expenses before any financial aid, scholarships, or grants are
applied. This price is typically higher than what most students will
pay, especially after financial aid is taken into account.

Public Net: The net price for attending a public college or university.
This is the amount that students and their families actually pay on
average after subtracting scholarships, grants, and financial aid from
the public sticker price. The net price provides a more accurate
representation of the cost to students and families after accounting for
available financial assistance.

Private Sticker: Similar to the public sticker price, this is the
advertised price of attending a private college or university. It tends
to be higher than the public sticker price, reflecting the total cost of
tuition, fees, room and board, and other expenses at a private
institution before any financial aid is applied.

Private Net: The net price for attending a private college or
university. It's the actual cost to students and their families after
deducting scholarships, grants, and financial aid from the private
sticker price. Since private colleges often offer significant financial
aid packages, the private net price can be considerably lower than the
sticker price, though it generally remains higher on average than the
public net price.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{4}
\tightlist
\item
  Income Distribution Data
\end{enumerate}

Log Household Income: This refers to the log-transformed household
income. In economic data analysis, income values are often
log-transformed for several reasons such as to reduce skewness and helps
linearize multiplicative relationships, making them easier to model and
interpret

Density (EMG\_ML\_PDF): EMG\_ML\_PDF'' stands for the Estimated Maximum
Likelihood Probability Density Function. This term combines several
concepts such as Estimated Maximum Likelihood (EML) which is a
statistical method used to estimate the parameters of a probability
distribution that make the observed data most probable. It's a common
approach in econometrics and statistics for fitting models to data and
Probability Density Function (PDF) which describes the likelihood of a
random variable to take on a given value. In practical terms, it shows
how the values of the random variable are distributed across different
outcomes. This density variable shows how the probability distribution
of incomes (after log transformation) changes across different income
levels for the years analyzed.

@@

Comparing the two data sets, we can see that even though the percentage
increase of private institution fees. The net expenditure is still
almost three times of public institutions. Overall by income group,
private institutions are also a lot more generous in giving out grants,
with large numbers compared to public institutions. Private institutions
also have a positive percentage increase of subsidy expenditure compared
to public institutions despite both having positive tuition revenue over
the years. From the table, we can see that private institutions are much
more generous in terms of grants and expenditure studies, however, they
are also a lot more expensive to attend compared to public schools.

@@

\includegraphics{college_tuiton_income_inequality_files/figure-pdf/fig-stickers-1.pdf}

@@

\includegraphics{college_tuiton_income_inequality_files/figure-pdf/fig-log-1.pdf}

@@

The paper by Zhifeng Cai and Jonathan Heathcote delves into the nuanced
dynamics between escalating college tuition fees and the widening gap of
income inequality in the United States. It posits a model showcasing how
increased income disparities significantly drive up tuition costs,
thereby impacting college accessibility. The study highlights the
strategic role of financial aid and subsidies in counteracting tuition
hikes, yet underscores their insufficiency in bridging the educational
divide (Cai and Heathcote 2022). Through a detailed examination of
higher education market dynamics, the paper sheds light on the
competitive practices of colleges, such as differential pricing and the
allocation of financial aid, aimed at attracting students. This analysis
culminates in a discussion on the broader implications for policy,
suggesting that reform in financial aid structures and a reevaluation of
higher education funding models are imperative to address these
challenges. The findings present a critical perspective on the economic
forces shaping the higher education landscape, offering insights that
could guide policy reforms aimed at ensuring equitable access to college
education amidst growing income inequality (Cai and Heathcote 2022).

@@

To explore the dynamics of Canadian universities through the insights
provided in the paper by Zhifeng Cai and Jonathan Heathcote and applying
a Canadian lens, it's vital to delve into the nuances of tuition trends,
income inequality, and access to higher education within Canada.
Canadian universities have been experiencing a rise in tuition fees
across various programs, with notable increases in professional degrees
such as dentistry, medicine, law, and engineering, reflecting a
potential barrier to access for students from lower-income backgrounds.
The disparities in tuition fees are further exacerbated when considering
the level of study, with graduate programs generally demanding higher
fees than undergraduate ones, aligning with the expected returns in
terms of future employment income (Longpré-Verret and Richards 2021).

@@

\subsection{Limitations to our
reproduction}\label{limitations-to-our-reproduction}

@@

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-caiandheathcote}
Cai, Zhifeng, and Jonathan Heathcote. 2022. {``College Tuition and
Income Inequality.''} \emph{AMERICAN ECONOMIC REVIEW} 112 (2): 81--121.
\url{https://doi.org/10.1257/aer.20181027}.

\bibitem[\citeproctext]{ref-cadstat}
Longpré-Verret, Léa-Maude, and Elizabeth Richards. 2021. \emph{Diversity
Among Board Directors and Officers: Exploratory Estimates on Family,
Work and Income}.
\url{https://www150.statcan.gc.ca/n1/pub/11f0019m/11f0019m2021005-eng.htm}.

\bibitem[\citeproctext]{ref-citeR}
R Core Team. 2023. \emph{R: A Language and Environment for Statistical
Computing}. Vienna, Austria: R Foundation for Statistical Computing.
\url{https://www.R-project.org/}.

\bibitem[\citeproctext]{ref-ggplot2}
Wickham, Hadley. 2016. \emph{Ggplot2: Elegant Graphics for Data
Analysis}. Springer-Verlag New York.
\url{https://ggplot2.tidyverse.org}.

\bibitem[\citeproctext]{ref-tidyverse}
Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy
D'Agostino McGowan, Romain François, Garrett Grolemund, et al. 2019.
{``Welcome to the {tidyverse}.''} \emph{Journal of Open Source Software}
4 (43): 1686. \url{https://doi.org/10.21105/joss.01686}.

\bibitem[\citeproctext]{ref-readxl}
Wickham, Hadley, and Jennifer Bryan. 2023. \emph{Readxl: Read Excel
Files}. \url{https://cran.r-project.org/web/packages/readxl/index.html}.

\bibitem[\citeproctext]{ref-dplyr}
Wickham, Hadley, Romain François, Lionel Henry, Kirill Müller, and Davis
Vaughan. 2023. \emph{Dplyr: A Grammar of Data Manipulation}.
\url{https://CRAN.R-project.org/package=dplyr}.

\bibitem[\citeproctext]{ref-tidyr}
Wickham, Hadley, Hadley Wickham, Davis Vaughan, Maximilian Girlich, and
Kevin Ushey. 2024. \emph{Tidyr: Tidy Messy Data}.
\url{https://cran.r-project.org/web/packages/tidyr/index.html}.

\bibitem[\citeproctext]{ref-knitr}
Xie, Yihui. 2023. \emph{Knitr: A General-Purpose Package for Dynamic
Report Generation in r}. \url{https://yihui.org/knitr/}.

\end{CSLReferences}

>>>>

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

@@

\title{Income Disparity and Its Influence on Higher Education
Affordability and Access\thanks{Code and data are available at:
https://github.com/fanger2791/College-Tuition-and-Income-Inequality/tree/main}}
\author{Michael Fang \and Harrison Huang}
\date{February 19, 2024}

@@

In recent years, the escalating cost of college tuition in the United
States has emerged as a critical issue, intertwining with the broader
narrative of rising income inequality. The paper by Zhifeng Cai and
Jonathan Heathcote addresses this pressing concern by examining the
relationship between these two phenomena. Utilizing a competitive model
of the college market, the study not only uncovers the mechanisms that
link rising income inequality to increasing tuition costs but also sheds
light on the consequent impacts on college attendance and social
mobility. Through this paper, we aim to illuminate the unique challenges
and opportunities facing the Canadian higher education system in the
context of rising tuition costs and income inequality. Incorporating a
Canadian perspective into our reproduction study adds a crucial
comparative dimension, allowing us to explore the dynamics of university
tuition and income inequality within a different national context. This
extension acknowledges the distinct structure of the Canadian higher
education system and its funding mechanisms, which differ from those in
the United States in terms of government support, the magnitude of
tuition fees, and the socio-economic landscape.

Central to the Zhifeng Cai and Jonathan Heathcote paper's findings is
the assertion that the surge in U.S. income inequality since the 1990s
can account for more than half of the observed increase in average net
tuition over the same period. This relationship underscores a critical
feedback loop where income disparities not only affect individual
capacity to afford higher education but also drive institutional
behaviors around tuition setting. Moreover, the paper highlights the
detrimental effects of rising tuition on college attendance rates,
particularly among lower-income segments, positing significant
implications for social mobility and equity.

@@

Our project involves crafting a concise paper that encapsulates the
entire reproduction study. This paper will meticulously outline our
source and methodology, ensuring clarity on how we've mirrored the
original study's model while integrating steps for high reproducibility.
It will also present a comparative analysis of our findings against the
original study's conclusions, pinpointing any variances and their
potential causes. Additionally, the paper will delve into the
implications of our policy simulations, offering insights into how
income inequality influences college tuition and attendance. Lastly, it
will propose directions for future research, highlighting areas within
the college tuition-income inequality nexus that remain unexplored,
setting the stage for subsequent scholarly inquiry.

@@

@@

In replicating the original model by Zhifeng Cai and Jonathan Heathcote,
our methodology was meticulously crafted to mirror their approach with
precision, ensuring the highest degree of reproducibility. This entailed
a thorough analysis of the original study's model, including its
assumptions, variables, and computational methods. We closely followed
the procedural steps outlined in their research, from data collection
through to the analytical techniques employed to examine the
relationship between income inequality and college tuition costs. To
guarantee reproducibility, we documented each step of our process in
detail, including the coding practices, statistical software used, and
the sources of our data. This approach not only underscores our
commitment to transparency and scientific integrity but also enhances
the reliability of our replication efforts.

R was the language and environment used for the bulk of this
reproduction, alongside

\subsection{Features}\label{features}

@@

Comparing the two data set, we can see that even though the percentage
increase of private institution fees. The net expenditure is still
almost three times of public institutions. Overall by income group,
private institutions are also a lot more generous in giving out grants,
with large numbers compared to public institutions. Private institutions
also have a positive percentage increase of subsidy expenditure compared
to public institutions despite both having positive tuition revenue over
the years. From the table, we can see that private institutions are much
more generous in terms of grants and expenditure studies, however, they
are also a lot more expensive to attend compared to public schools.

@@

\includegraphics{paper_files/figure-pdf/fig-stickers-1.pdf}

@@

\includegraphics{paper_files/figure-pdf/fig-log-1.pdf}

@@

The paper by Zhifeng Cai and Jonathan Heathcote delves into the nuanced
dynamics between escalating college tuition fees and the widening gap of
income inequality in the United States. It posits a model showcasing how
increased income disparities significantly drive up tuition costs,
thereby impacting college accessibility. The study highlights the
strategic role of financial aid and subsidies in counteracting tuition
hikes, yet underscores their insufficiency in bridging the educational
divide. Through a detailed examination of higher education market
dynamics, the paper sheds light on the competitive practices of
colleges, such as differential pricing and the allocation of financial
aid, aimed at attracting students. This analysis culminates in a
discussion on the broader implications for policy, suggesting that
reform in financial aid structures and a reevaluation of higher
education funding models are imperative to address these challenges. The
findings present a critical perspective on the economic forces shaping
the higher education landscape, offering insights that could guide
policy reforms aimed at ensuring equitable access to college education
amidst growing income inequality.

@@

To explore the dynamics of Canadian universities through the insights
provided in the paper by Zhifeng Cai and Jonathan Heathcote and applying
a Canadian lens, it's vital to delve into the nuances of tuition trends,
income inequality, and access to higher education within Canada.
Canadian universities have been experiencing a rise in tuition fees
across various programs, with notable increases in professional degrees
such as dentistry, medicine, law, and engineering, reflecting a
potential barrier to access for students from lower-income backgrounds.
The disparities in tuition fees are further exacerbated when considering
the level of study, with graduate programs generally demanding higher
fees than undergraduate ones, aligning with the expected returns in
terms of future employment income.

@@

\subsection{Limitations to
reproducing}\label{limitations-to-reproducing}

@@

\section{References}\label{references}
